Stochastic Approximation Monte Carlo with Dynamic Update Factor (SAD)

### 2018-08-06 papers/histogram/sad.tex
update paper with new sad

{\color{red}
While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy, this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by the energy difference
$E_H -E_L$ where these energies are defined below in the following
section.  Early in the simulation, when the energy range is still
expanding, this average entropy is a good choice
for $t_0$.  However, as the simulation progresses, our entropy factor
needs to scale with the total number of energy states that have been
found $N_S$, since the product $N_S\langle S\rangle$ corresponds to the
total change of entropy required, given that each of the $N_S$ states
will has a separate weight $w(E)$.  Finally, at \emph{long} times, after
all the energies have been found and the weights are being refined
we wish for a lower $t_0$ in order to reduce the fluctuations in the
weights that result from large values for $\gamma_t$.  We enable this
by tracking the last time we encountered a new energy $t_L$, and
gradually transition to lower $t_0$ (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{E_{H}-E_{L}}{3T_{\text{min}}t}
    \frac{
      E_S^2 + E_S t + \left(t/t_L-1\right)t
    }{
      E_S^2 + t + \left(t/t_L-1\right)t
    }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is discovered during the course of
the simulation, the update factor experiences a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly. Figure~\ref{fig:gamma-vs-t} compares $\gamma_t$ for SAD, WL,
and SAMC.  The update factor
for WL remains high for many iterations, and then decreases very rapidly,
while the SAMC $\gamma_t$ remains constant before dropping as $1/t$.
The update factor for SAD in contrast fluctuates
dynamically around a value less than 1 for early MC moves
(while finding new energy states), before dropping in a \davidsays{similar
to SAD, but described after we see the plot}.
}

>>>>

{\color{red}
While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy, this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by the energy difference
$E_H -E_L$ where these energies are defined below in the following
section.  Early in the simulation, when the energy range is still
expanding, this average entropy is a good choice
for $t_0$.  However, as the simulation progresses, our entropy factor
needs to scale with the total number of \davidsays{interesting} energy states that have been
found $N_S$, since the product $N_S\langle S\rangle$ corresponds to the
total change of entropy required, given that each of the $N_S$ states
will has a separate weight $w(E)$.  Finally, at \emph{long} times, after
all the energies have been found and the weights are being refined
we wish for a lower $t_0$ in order to reduce the fluctuations in the
weights that result from large values for $\gamma_t$.  We enable this
by tracking the last time we encountered a new energy $t_L$, and
gradually transition to lower $t_0$ (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     N_S \frac{
       \frac{E_{H}-E_{L}}{3T_{\text{min}}} + \frac{t}{t_L}
     }{
       N_S^2 + \frac{t^2}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found in the
\davidsays{interesting energy range} during the course of
the simulation, the update factor experiences a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly. Figure~\ref{fig:gamma-vs-t} compares $\gamma_t$ for SAD, WL,
and SAMC.  The update factor
for WL remains high for many iterations, and then decreases very rapidly,
while the SAMC $\gamma_t$ remains constant before dropping as $1/t$.
The update factor for SAD in contrast fluctuates
dynamically around a value less than 1 for early MC moves
(while finding new energy states), before dropping in a \davidsays{similar
to SAD, but described after we see the plot}.
}
